A non-additive generalization of probability theory \\for quantum mechanics and beyond

### 2025-07-21 Articles/2024-NonAdditiveProbability/NonAdditiveProbability.tex
NonAdditiveProbability: reworked Choquet theory

The set of all extreme points $X$, as we will see, plays a major role. However, on physical grounds, these may not be proper ensembles. In classical mechanics, these would corresponds to Dirac measure: probability distributions all concentrated at a single point of phase space. Since we cannot reliably prepare a particle with infinite precision, these are not ensemble and they are not part of $\Ens$. For the purpose of Choquet theory, we will assume that there is a way to find the closure $E \supset \Ens$ that includes all extreme points $X$. Moreover, we will also assume $E$ to be compact, even though it will not be. How to properly make these identification, and how to go beyond the compact case, is outside of the scope of this particular article.\footnote{An early attempt used the notion of orthogonality to create the lattice of orthogonal subspaces and use something along the lines of Stone's representation theorem for Boolean algebra. This does not work. An approach, currently under investigation, is to study limits over statistical quantities. We are still trying to understand what mathematical structures have enough information to allow the reconstruction.} Here we will use $\Ens$ to indicate the general case and $E$ to indicate this ``compact closure.''

@@

\textbf{Classical probability measures.} In the classical case, ensembles are defined as probability measures of the space of all possible states $X$. In the discrete case, like for roll of a die, $X$ is a set endowed with the discrete topology\footnote{Physically, the discrete topology corresponds to the ability to experimentally verify or falsify each case. That is, we are able to recognize whether a particular die landed or didn't land on a particular face.} and, consequently, the power set as the $\sigma$-algebra. In the continuous case, $X$ will be classical phase space endowed with the usual topology of a manifold\footnote{Physically, the topology of a manifold corresponds to the ability to label each possible case through a set of continuous variable which are verifiable up to finite but arbitrarily small precision. That is, we are able to tell whether a state is within a particular range of position and momentum.} and the standard Borel sets as the $\sigma$-algebra. In either case, a probability measure is a map $p : \Sigma_X \to [0,1]$ that assign to each event of the $\sigma$-algebra (i.e. to an empirical statements) the probability to find it to be true. The probability measure is additive, meaning that if two events $A, B \in \Sigma_X$ are mutually exclusive (i.e. $A \cap B = \emptyset$) then the probability of the union is the sum of the probability of the respective events:
\begin{equation}
	p(A\cup B) = p(A) + p(B).
\end{equation}

\textbf{Expectation and integration.} TODO: show the use of integrals over probability measure.

\textbf{Absolute continuity of probability measures.} While each classical ensemble is a probability distribution, not all probability distribution can be thought as a valid ensemble. For example, we could have a probability measure to be wholly defined over a single point of phase space, which would correspond to the ability to prepare a system in the same exact place with the same exact momentum. Not only this is not feasible experimentally, which violates the whole premise of having objects that map to physically meaningful entities, but it would have to have entropy equal minus infinity. To address this, we add a requirement that is both mathematically natural and physically meaningful. Let $\mu : \Sigma_X \to [0, + \infty]$ the measure on $X$ that quantifies the count of states in each region. In the discrete case this is simply the counting measure (i.e. three possible faces of the die corresponds to three possible states), while in the continuous case this corresponds to the Liouville measure (i.e. volumes in phase space correspond to the count of states). This measure is additive over mutually exclusive events. The requirement is that we can assign non-zero probability only to events that have a non-zero count of states. That is, if for a particular event $A$ we have $p(A) \neq 0$, then we must have that $\mu(A) \neq 0$ as well. Mathematically, $p$ is absolutely continuous with respect to $\mu$, and this is exactly the case in which a probability density $\rho = \frac{dp}{d\mu}$ can be defined.\footnote{In the general case, we can still assign non-zero probability to a point of a manifold, provided that the topology and the measure $\mu$ consistently allow for it. Physically, it means that those points are ``special'' both in terms of experimental verifiability and entropy/count of states. While the general framework allows for it, it is not crucial for our discussion here.} This is another example in which a physical requirement (probability must be zero if there are no possible states), once properly spelled out (absolute continuity), justifies common physical intuition (we assume there is a probability density).

% Intro paragraph that says quickly how to connect to Choquet theory and its limitations. It should be a quick summary of the whole section

\textbf{Choquet theory.} Having reviewed the use of probability measures in classical mechanics, we can ask the more general question: when can we represent ensembles in a generic ensemble space as a probability measure? A very similar question is the subject of Choquet theory.

First, we have to define what it means to represent an ensemble with a probability measure. As a first step, let $E$ be a convex subset of a locally convex space $V$. We say that $\ens \in E$ is represented by a probability measure $p : \Sigma_{E} \to [0,1]$ if $f(\ens) = \int_E f dp$ for every continuous linear functional $f$ on $V$. Physically, instead of describing our ensemble $\ens$ as a point in our ensemble space identified by the value of all possible statistical variables $f(\ens)$, we represent it as a probability measure $p$ and all the associated expectation values.

In this first step, we are using a probability measure on $E$, so we are defining a probability distribution over ensembles. As a second step, we want the probability measure to be defined over the possible cases, the sample space. To do that, Choquet theory identifies each possible case with the Dirac measure associated to that case, which is an extreme point of the convex set $E$. The result is the following:
\begin{thrm}[Choquet]
	Let $E$ be a metrizable compact convex subset of a locally convex space $V$, and let $\ens \in E$. Then there exists a probability measure $\mu_{\ens}$ on $X$ which represents $\ens$ and is supported by the extreme points of $X$.
\end{thrm}

Note that while an ensemble space $\Ens$ is a metrizable convex subset of a locally convex TVS $V$, it is not necessarily compact. As we saw, the Dirac measures should not be considered valid ensembles in the continuum case. Therefore Choquet theory cannot be applied to our general case. When we are drawing the parallel, we will implicitly assume that $E$ is a compact set that is also the closure of $\Ens$ in an appropriate convex TVS $V$. Many interesting cases fall into this restriction: the ensemble spaces of classical probability distribution over finitely many discrete cases fall in this restriction; the subset of ensembles over finite ranges of position and momentum; the ensemble spaces of all finite dimensional quantum systems. When Choquet theory applies, then, we are able to always characterize an ensemble in terms of a probability distributions over its, maybe idealized, most refined ensembles.

The previous theorem does not, however, guarantee uniqueness. The above theorem, in fact, applies in the quantum case. The Bloch ball, in fact, is a compact convex subset and any point can be seen as a convex combination of points on the surface. The center, which represent the maximally mixed state, can be thought as an equal mixture of any two opposite points, or a uniform distribution over the whole surface. The fact that any ensemble can be seen as a mixture of a set of pure states, then, is not a property of classical spaces. The property of classical spaces is that this representation is unique.

Mathematically, the probability measure is unique if and only if $E$ is a Choquet simplex, which is a generalization of the finite dimensional simplex. The characterization of the Choquet simplex is fairly technical, so we will simply define the Choquet simplex by the property of having a single decomposition in terms of extreme points. The classical case, that is when the ensemble space coincide with the space of all probability measures over a set of states, is a Choquet simplex.

>>>>

The set of all extreme points $X$, as we will see, plays a major role. However, on physical grounds, these may not be proper ensembles. In classical mechanics, these would corresponds to Dirac measure: probability distributions all concentrated at a single point of phase space. Since we cannot reliably prepare a particle with infinite precision, and its entropy would correspond to minus infinity, these are not ensemble and they are not part of $\Ens$. For the purpose of Choquet theory, we will assume that there is a way to find the closure $E \supset \Ens$ that includes all extreme points $X$. Moreover, we will also assume $E$ to be compact, even though it will not be in general. How to properly make these identification, and how to go beyond the compact case, is outside of the scope of this particular article.\footnote{An early attempt used the notion of orthogonality to create the lattice of orthogonal subspaces and use something along the lines of Stone's representation theorem for Boolean algebra. This does not work. An approach, currently under investigation, is to study limits over statistical quantities. We are still trying to understand what mathematical structures have enough information to allow the reconstruction.} Note that the classical discrete and the quantum case fall in the compact case. Moreover, the set of all probability distributions over a compact space $X$ also fall in the compact case. Therefore, apart from the technicality above, the compact case is still a very interesting case. To distinguish the context, in this article we will use $\Ens$ to indicate the general case and $E$ to indicate this ``compact closure.''

@@

\textbf{Classical probability measures.} In the classical case, given the topological space of possible states $X$ (i.e. the sample space), ensembles are defined as probability measures $p : \Sigma_X \to [0,1]$. That is, a probability measure assigns to each element of the Borel algebra $\Sigma_X$\footnote{In our overall project, Borel sets represent statements associated to a test, regardless of termination.}, an event, the probability to find it to be true. In the discrete case, like for roll of a die, $X$ is a set endowed with the discrete topology\footnote{Physically, the discrete topology corresponds to the ability to experimentally verify or falsify each case. That is, we are able to recognize whether a particular die landed or didn't land on a particular face.} and, consequently, the power set is the $\sigma$-algebra. In the continuous case, $X$ will be classical phase space endowed with the usual topology of a manifold.\footnote{Physically, the topology of a manifold corresponds to the ability to label each possible case through a set of continuous variable which are verifiable up to finite but arbitrarily small precision.}  The probability measure is countably additive, meaning that, given a countable collection $A_i$ of pair-wise disjoint events (i.e. $A_i \cap A_j = \emptyset$), the probability of the union is the sum of the probability of the respective events:
\begin{equation}
	p\left(\bigcup_i A_i \right) = \sum_i p(A_i).
\end{equation}
This last requirement codifies the idea that disjoint event are mutually exclusive: if $A$ and $B$ cannot happen at the same time, then the probability of one of the two happening is the sum of them separately happening.

\textbf{Expectation and integration.} A random variable is a function $F : X \to \mathbb{R}$ that assigns a value to each state.\footnote{The function $f$ must be measurable, in the sense that the inverse of a Borel should be a Borel set. We find that the term is confusing for a physics audience, so it is avoided in the main text.} The number of the side of the die, position, momentum are typical examples. The expectation for a random variable is given by the integral
\begin{equation}
	E[F] = \int_X F dp.
\end{equation}
The expectation is linear with respect to the variable (i.e. $E[aF+bG]= aE[F]+bE[G]$). In our case, what is more interesting is that it is a convex function in terms of probability measures: if $p = \lambda p_1 + \bar{\lambda} p_2$ then $\int_X F dp = \lambda \int_X F dp_1 + \bar{\lambda} \int_X F dp_2$. That is, as we required for statistical variables, the expectation for a mixture is the mixture of the expectations.

\textbf{Absolute continuity of probability measures.} As we saw before, probability measures over a single point over the continuum cannot be understood as valid ensembles. To address this, we add a requirement that is both physically meaningful and mathematically useful. Let $\mu : \Sigma_X \to [0, + \infty]$ be a measure on $X$ that quantifies the count of states in each region. In the discrete case this is simply the counting measure (i.e. three possible faces of the die corresponds to three possible states), while in the continuous case this corresponds to the Liouville measure (i.e. volumes in phase space correspond to the count of states). This measure is additive over disjoint region as they represent mutually exclusive events. The requirement is that we can assign non-zero probability only to events that have a non-zero count of states. That is, if for a particular event $A$ we have $p(A) \neq 0$, then we must have that $\mu(A) \neq 0$ as well. Mathematically, $p$ is absolutely continuous with respect to $\mu$, and this is exactly the case in which a probability density $\rho = \frac{dp}{d\mu}$ can be defined.\footnote{In the general case, we can still assign non-zero probability to a point of a manifold, provided that the topology and the measure $\mu$ consistently allow for it. Physically, it means that those points are ``special'' both in terms of experimental verifiability and entropy/count of states. While the general framework allows for it, it is not crucial for our discussion here.} This is another example in which a physical requirement (probability must be zero if there are no possible states), once properly spelled out (absolute continuity), justifies common physical intuition (we assume a probability density exists).

\textbf{Choquet theory.} Classical statistical mechanics and probability starts by assuming the existence of a state space $X$ and defining ensembles on top of it. This cannot be generalized to quantum mechanics and beyond. In our theory of ensemble spaces we start with ensembles, and have to show when and how they can be thought, can be represented, as probability measures. If we can assume that $E$ is a compact convex set, Choquet theory answers this question.

First, we have to define what it means to represent an ensemble with a probability measure. Let $E$ be a convex subset of a locally convex space $V$ and $X$ the set of extreme point. In classical mechanics, $X$ will correspond to the set of Dirac measures, as these cannot seen as a mixture of other measures. In quantum mechanics, $X$ corresponds to the pure states, as these cannot be seen as mixture of other states. In other words, the set of states $X$ is recovered as the set of ensembles with the narrowest spread. We say that $\ens \in E$ is represented by a probability measure $p : \Sigma_{E} \to [0,1]$ if $f(\ens) = \int_E f dp$ for every continuous linear functional $f$ on $V$. Moreover, $p$ is supported by the extreme points if $p(X) = 1$. Physically, we start with an ensemble $\ens$ taken as a point in our ensemble space, identified by the value of all possible statistical variables $f(\ens)$, and we represent it as a probability measure $p$, a statistical mixture, over other the extreme points, the states, recovering the expectation with an integral.

We should note that, given any probability measure $p$ over the extreme points $X$, we can always find an $\ens \in E$ that is represented by $p$. Therefore, we have a map from the set of probability measures to the set of ensembles. The first question, then, is whether we can go back. Choquet theory tells us that, in the compact case, we can:
\begin{thrm}[Choquet 1]
	Let $E$ be a metrizable compact convex subset of a locally convex space $V$, and let $\ens \in E$. Then there exists a probability measure $p_{\ens}$ on $X$ which represents $\ens$ and is supported by the extreme points $X$.
\end{thrm}
Note that an ensemble space is a metrizable convex subset of a locally convex TVS. Therefore, if $E$ is compact, the Choquet theorem applies.

The above theorem tells us that the map between probability measures and ensembles is surjective. The second question, then, when is it bijective? Again, Choquet theory has a clear answer.
\begin{thrm}[Choquet 2]
	Let $E$ be a closed convex metrizable subset of a locally convex space $V$. Then $E$ is a Choquet simplex if and only if for each $\ens \in E$ there exists a unique measure $p_{\ens}$ which represents $\ens$ and is supported by the extreme points $X$.
\end{thrm}
The Choquet simplex is a generalization of the finite dimensional simplex, whose mathematical characterization is rather technical. The key point is that $E$ is a Choquet simplex if and only if the ensemble space does not allow multiple decomposition in terms of pure states. This confirms the physical intuition from classical and quantum statistical mechanics.

For example, the ensemble space for a two state quantum system, the Bloch ball, is a compact convex subset. Any internal point can be seen as the convex combination of points on the surface, the pure states. The center, which represents the maximally mixed state, can be thought as an equal mixture of any two opposite points, or a uniform distribution over the whole surface. The fact that any ensemble can be seen as a mixture of a set of pure states, then, is reflected by the first part of the Choquet theorem, and it is valid for all physical theories. Uniqueness of the representation fails, however, precisely because the ensemble space is not a simplex: it is not a classical space.
